\ \\ \LARGE\bf Speculative Evaluation in Particle Swarm Optimization \thanks{Matthew Gardner, Andrew McNabb, and Kevin Seppi are with the Department of Computer Science, Brigham Young University, 3361 TMCB, Provo, UT 84602 (phone: 801-422-8717; email: \{mjg82,a,k\}@cs.byu.edu).}

### 2010-04-02 honors_thesis/journal_paper/paper.tex
Major revisions to the journal paper from feedback at the meeting

%% PSO Stuff
\DeclareMathOperator*{\argmin}{arg\;min}
\DeclareMathOperator*{\argmax}{arg\;max}
\DeclareMathOperator*{\arginf}{arg\;inf}
\DeclareMathOperator*{\argsup}{arg\;sup}
\providecommand{\pers}{\ensuremath{P}}
\providecommand{\neigh}{\ensuremath{N}}
\providecommand{\leftind}{\ensuremath{L}}
\providecommand{\rightind}{\ensuremath{R}}
\providecommand{\nURand}{\ensuremath{U^\neigh}}
\providecommand{\pURand}{\ensuremath{U^\pers}}
\providecommand{\ppos}{\ensuremath{\Vec{x}}}
\providecommand{\fppos}{\ensuremath{\Vec{fx}}}
\providecommand{\pvel}{\ensuremath{\Vec{v}}}
\providecommand{\nbest}{\ensuremath{\Vec{b}^\neigh}}
\providecommand{\fnbest}{\ensuremath{\Vec{fb}^\neigh}}
\providecommand{\pbest}{\ensuremath{\Vec{b}^\pers}}
\providecommand{\fpbest}{\ensuremath{\Vec{fb}^\pers}}
\providecommand{\constriction}{\ensuremath{\chi}}
\providecommand{\ncoeff}{\ensuremath{\phi^\neigh}}
\providecommand{\pcoeff}{\ensuremath{\phi^\pers}}
\providecommand{\obs}{\ensuremath{\Vec{\xi}}}
\providecommand{\ofunc}{\ensuremath{f}}
\providecommand{\swarm}{\ensuremath{swarm}}

%SpecExPSO Stuff
\providecommand{\indic}{\ensuremath{I}}
\providecommand{\specvel}{\ensuremath{\vec{V}}}
\providecommand{\specpos}{\ensuremath{\vec{X}}}
\providecommand{\leftn}{\ensuremath{\Vec{x}^\leftind}}
\providecommand{\rightn}{\ensuremath{\Vec{x}^\rightind}}
\providecommand{\caseset}{\ensuremath{\mathcal{C}}}
\providecommand{\casedef}{\ensuremath{(\pbest,\nbest)}}
\providecommand{\casexn}{\ensuremath{(x,\neigh)}}
\providecommand{\casexx}{\ensuremath{(x,x)}}
\providecommand{\casexl}{\ensuremath{(x,\leftind)}}
\providecommand{\casexr}{\ensuremath{(x,\rightind)}}
\providecommand{\casepn}{\ensuremath{(\pers,\neigh)}}
\providecommand{\casepl}{\ensuremath{(\pers,\leftind)}}
\providecommand{\casepr}{\ensuremath{(\pers,\rightind)}}
\providecommand{\noeval}[1]{\ensuremath{#1^{-e}}}
\providecommand{\nonbest}[1]{\ensuremath{#1^{-n}}}
\providecommand{\p}{\ensuremath{p}}
\providecommand{\pset}{\ensuremath{\mathbf{p}}}
\providecommand{\s}{\ensuremath{s}}
\providecommand{\sset}{\ensuremath{\mathbf{s}}}
\providecommand{\nsset}{\ensuremath{\mathbf{ns}}}
\providecommand{\n}{\ensuremath{n}}
\providecommand{\nset}{\ensuremath{\mathbf{n}}}
\providecommand{\nnset}{\ensuremath{\mathbf{nn}}}

@@

Particle swarm optimization (PSO) has previously been parallelized primarily by 
distributing the computation corresponding to particles or sets
across multiple processors.  However, in many cases this approach is not efficient when scale to 
very large populations sizes (on very large clusters).
In this paper we present a different way of using multiple processors of the 
computation in PSO, motivated by the ``speculative execution'' in modern computer hardware and which we call
Speculative Evaluation in PSO (SEPSO).

In SEPSO future positions of the particles are
speculated and evaluated in parallel with the evaluation of current positions.
We show that a set of separate speculative evaluation threads can be constructed
such that the set will always include the position and corresponding
evaluation that would normally have occurred at the next iteration of standard PSO.
Thus SEPSO performs two
iterations of PSO at once, allowing it to achieve better fitness faster.
Not only can SEPSO can achieve better fitness in the same number of iterations,
it also reaches better fitness for the same number of total function evaluations
in large scale parallel environments.

Speculative Evaluation in PSO can be enhanced by diverging from the
behavior of the standard PSO to make better use of the information
obtained by the speculative evaluations.
Using these algorithmic modifications, we see dramatic improvements in the
effectiveness of the algorithm. 
\end{abstract}

@@

Particle swarm optimization (PSO) has been found to be a highly robust and
effective algorithm for solving many types of optimization problems.  For much
of the algorithm's history, PSO was run serially on a single machine.  However,
the world's computing power is increasingly coming from large clusters of
processors.
%Even in desktop machines, systems with 4 or 8 cores are
%commonplace.  
In order to efficiently utilize these resources for
computationally intensive problems, PSO needs to run in parallel.

@@

These previous parallel techniques 
distribute the computation needed by the particles in the swarm over the available processors.
if more processors are available, these prior techniques increase the
number of particles in the swarm, either by adding individual particles or by
adding entire new sub-swarms but in all cases
the number of processors never exceeds the number of particles.
The number of iterations of PSO that the
algorithms can perform is thus inherently limited by the time it takes to
evaluate the objective function---additional processors add more particles, but do not
make the iterations go any faster.

The purpose of this paper is to consider PSO parallelization strategies for
clusters of thousands of processors and functions for which a single
evaluation will take at least tens of seconds, but probably hours or perhaps days.
Our purpose is to explore the question ``Since PSO is often run with
a swarm of size of 50-100, what would one do if a few thousand processors were available?''

For many functions there comes a point of diminishing returns with respect to adding particles.
In \fig{basic-sphere} we show the value obtained after 20,000 function evaluations (not iterations)
as a function of swarm size
for the well-know benchmark function Sphere (20 dimensions, reporting the average of thirty runs). 
Increasing the swarm size from 10 to 50 has a significant effect value obtained
however, increasing the swarm size from 100 to 200 makes the algorithm less efficient,
that is, it reduces the progress the algorithm makes per-evaluation.
For this reason, previous work has recommended the use a swarm size of 50 for this function
~\cite{bratton-sis07}.
Thus, in at least some cases, added particles indefinitely will not yield an efficient implementation.

@@

To solve this problem, we take inspiration from speculative execution
techniques commonly used in processors.  Modern processors, when faced with a
branch on which they must wait (e.g. a memory cache miss), guess which way the
branch will go and start executing, being careful not to make changes that can not be undone.
If the processor guesses right, execution
is much farther ahead than if it had just idly waited on the memory reference.  If it
guesses wrong, execution restarts where it would have been anyway.
This is called speculative execution or branch prediction.

\begin{quote}
SKIP SKIP

No really, try reading without this! The idea of speculative execution can be applied to some types of optimization
algorithms.  When the next sampling position of the algorithm does not depend
on the value of the objective function at the current position,
future positions can be speculated and evaluated in parallel with the current
position.
This is trivially true for optimization by random sampling, for example.
The next random sample does not depend on the previous point nor on the value
of the objective function at that point in any way.
It depends only on a random number.
There is no need to wait to see
what objective function value was obtained at the previous point,
evaluation of the object function at the next point can begin as soon as the next random number is drawn.
In other optimization algorithms, the next position depends on the previous position, but not the
value obtained there, a random walk for example.
In this case
the next position to be evaluated
depends only on the previous position and a random number.
In this case, as it was in the case of random sampling,
the object function evaluation of the next position can be done in parallel
because it 
does not depend on the \emph{value} of the objective function at the previous position.
There are an infinite number of possible next positions, but only one given the draw of the random number.

In still other cases, the next position to be evaluated may depend on the current position and some
discrete valued (binary, for example) test on the value obtained at the current position.
For example, consider a greedy random walk, in which the state only advances if a better
position is found.
That is, if the current value is better than the previous value,
use the sum of the current position and the random number as the next position.
Otherwise, computer the next position using the sum of the previous position and the random number.
In this case the objective function can be evaluated at three positions:
(1) the value of the objective function at the current position,
(2) the value at the position if the test turns out to be true and (3) the value at the position
if the test turns out to be false; all in parallel.
In this example it is important to note that we do not need the value of the objective function at the current
point to to reduce the infinite number of possible next positions to just two alternatives.
Choosing between these two alternatives can be done post-hoc, after the objective function value
at the current position and at the two alternative are computed.
For all but the most computationally-trivial objective functions, this approach saves almost half of the
``wall-clock'' time.
This approach has been applied to simulated annealing~\cite{witte-tpds91} which like the greedy random walk,
depends on positions and a binary test.

This approach can not be used in gradient methods where the
step size for sampling typically depends on the gradient and therefor the position and 
a set of infinitely many possible \emph{values} outcomes
obtained at the current position (not binary or even finite).
\end{quote}

@@

In this paper we show that the results of standard PSO can be reproduced \emph{exactly}, two iterations at a time, using
a speculative approach like the ones described above. We prove that the standard PSO equation can
be factored such that a small set of speculative positions can be found which will \emph{always}
include the position computed in the next iteration.
By computing the value of the objective function for each of the speculative positions at
the same time the algorithm evaluates the objective function for the current position,
It is possible to know the objective function values for both the current and the next iteration at the same time.
We demonstrate this principle by implementation and show that it
produces exactly the same results as standard PSO, but two iterations at a time.
The resulting implementation runs efficiently on large clusters where the number of processors is much larger than
a typical or reasonable number of particles. Producing better results in less ``wall-clock'' time and,
surprisingly, with fewer total function evaluations.

Furthermore, we show that if we relax the requirements of the algorithm, no
longer demanding that it strictly reproduce the exact behavior of standard PSO, we
can introduce new speculative techniques that often out-perform PSO.
These variants make better use of the extra information obtained from
the extra exploration made by the speculative function evaluations.
We also explore the idea that, like branch prediction in processors, we
need not speculatively evaluate \emph{all} possible future positions,
we can accelerate the algorithm even if we are just \emph{likely} to have guessed right.
By pruning the speculation to just paths that are statistically likely to reproduce
the paths that are equivalent to PSO we can also 
speculating several iterations ahead instead of just one.
We also consider several recovery strategies for cases where
the pruned set of speculative evaluations does not contain the evaluation that standard PSO would have done.

@@

At each iteration ($t+1$) of the algorithm, the position $\ppos_t$ and velocity
$\pvel_t$ of each particle are updated as follows:
\begin{align}
\nonumber
	\pvel_{t+1} &=
		\constriction \bigl[ \pvel_t
			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
\label{eq:velupdate}
			& \quad \quad \quad \, + \ncoeff\nURand_{t}\otimes(\nbest_{t} - \ppos_{t})
		\bigr] \\
\label{eq:posupdate}
	\ppos_{t+1} &= \ppos_{t} + \pvel_{t+1}
\end{align}
where \( \nURand_{t} \) and \( \nURand_{t} \) are vectors of independent
and identically distributed random numbers drawn from a uniform
distribution, the \( \otimes \) operator is an element-wise vector
multiplication, $\pbest$ (called personal best) is the best position the
current particle has seen, and $\nbest$ (called neighborhood best) is the best
position the neighbors of the current particle have seen~\cite{bratton-sis07}.  \( \ncoeff \), \(
\pcoeff \), and \( \constriction \) are parameters with prescribed values
required to ensure convergence (2.05, 2.05, and .73,
respectively)~\cite{clerc-tec02}. Note also that we use a subscript for the 
time step $t$ or $t+1$ and have moved all other qualifiers,
such as the personal qualifier ``$\pers$'' to the superscript position.
Superscripts can be miss read as exponents, but they are qualifiers throughout this paper.
We use $p$ to indicate the number of particles in the swarm.

There are many ways of defining the neighbors of a particle. 
Changing the
neighborhood topology has a significant effect on the performance of the algorithm and 
this effect varies widely by function type.
%, as would be expected by the No Free Lunch
%theorem~\cite{wolpert-tec97}.
The two most common topologies used in the
literature are the Ring topology and the Complete
topology~\cite{bratton-sis07}.  In the Ring topology, each particle has one
(or more) neighbors to either side of it; in the Complete topology, every particle is a
neighbor to every other particle.  It is generally assumed that, a particle is also a
neighbor to itself in that its own position and value is considered when updating the
particle's personal best, $\nbest$.  Thus with $p$ particles, using the Ring topology each
particle has three neighbors: $i-1$, $i$ (itself), and $i+1$.  With the Complete
topology, each particle has $p$ neighbors.
% TODO USE THIS LATER!
% In this paper we use these
% topologies, as well as a parallel adaptation of the Complete topology, called
% Random, that has been shown to simulate the behavior of Complete with far less
% communication~\cite{mcnabb-cec09}.  In the Random topology, each particle
% randomly picks two other particles to share information with at each iteration,
% along with itself.  Thus in both the Ring and the Random topologies, all
% particles have 3 neighbors.

@@

The PSO algorithm can be trivially parallelized by distributing the processing needed for each
particle on a separate processor.
But as we have seen in the introduction, for some
functions, and for large numbers of processors just adding particles
reaches a point of diminishing returns.
That is, adding processors does not help us reach any given level of fitness appreciably faster.

To address this issue we present here a different use of additional processors.
We take inspiration from speculative execution in processors and instead of adding particles
we will employ a speculative evaluation approach like those described in the introduction. 
This allows us to perform iterations two at a time.

To do so we must first re-factor PSO such that the determination of the $\pbest$ and $\nbest$,
which depend upon the value of the objective function, is sepearate from the rest of the computation.
%First consider formula for the \emph{next} iteration of PSO, which is identical
%to Equations~\eqref{eq:velupdate} and \eqref{eq:posupdate} except the time
%subscripts are incremented.  As noted above and
For simplicity in this
explanation, we restrict our discussion here to the case of a Ring topology
with two neighbors which we will call the ``right neighbor'' and ``left
neighbor'' neighbors identified for each particle as $\rightn$ and $\leftn$
respectively.

Next note that the computation of $\pbest$ and $\nbest$ is not seen in the
update equations~\eqref{eq:velupdate} and \eqref{eq:posupdate}, so we will make
this mathematically explicit.  This update is usually implemented as a pair of
independent ``if``-statements, but that is a bit awkward for our purposes
so we instead use a set of indicator variables, one each representing each
possible update to the pair $\casedef$.  These indicator variables allow
us to easily identify the cases in which updates will be needed for $\nbest$,
$\pbest$ or both.

For a topology with 2 neighbors there are 7 possible update cases identified in
Table~\ref{tab:evals}.
For example and as indicated in the table, we write $\casexn$ to represent the case where $\pbest$ will be
updated with the better position $\ppos_{i,t}$, shown in the first positions,
but a no new neighborhood best was found so $\nbest$ remains in the second
position.
%This list assumes, as is typically the case for the ring topology, that each
%particle is also part of its own neighborhood (can cause an update to
%$\nbest$).
The list omits the case where $\nbest$ is updated by
$\ppos_{t}$ but $\pbest$ was not updated, as this is logically impossible.
$\caseset$ represents the set of all possible update cases.

For minimization, we can define the indicator variable representing the case $\casexn$,
where the particle will need to update $\pbest$ with the better
position $x_{t}$, but $\nbest$ will not updated, by:
\begin{align}
\nonumber
	\indic_{t+1}^{\casexn} & (\ofunc ( \ppos_{t+1} ) ,\ofunc(\leftn_{t+1}),\ofunc(\rightn_{t+1}) ,\ofunc(\pbest_{t}) ,\ofunc(\nbest_{t}))= \\
\nonumber
	& 1 \ if \  \ofunc(\ppos_{t+1}) > \ofunc(\pbest_{t}) \\
\nonumber
	& \quad \quad \ \wedge \ \ofunc(\nbest_{t}) > \ofunc(\ppos_{t+1}) \\
\nonumber
	& \quad \quad \ \wedge \ \ofunc(\nbest_{t}) > \ofunc(\leftn_{t+1}) \\
\nonumber
	& \quad \quad \ \wedge \ \ofunc(\nbest_{t}) > \ofunc(\rightn_{t+1}) \\
\label{eq:deficasexn}
	& 0 \ otherwise
\end{align}

For compactness in our equations we represent this without the parameters since
they can be inferred from the subscripts on $\indic_{t}^{\casexn}$.  The
indicator variables for the other update cases shown in Table~\ref{tab:evals}
($\casepn$, $\casepl$, $\casepr$, etc.) can be defined in similar fashion but
obviously using the corresponding boolean expression.

For each cases identified in Table~\ref{tab:evals} and identified by a
corresponding indicator variable defined above, we also provide notation for the velocity
update equation; once again using $\casexn$ as an example:
\begin{align}
\nonumber
	\specvel_{t+1}^{\casexn} & (\pvel_t, \ppos_{t+1}, \leftn_{t+1}, \rightn_{t+1}, \pbest_{t},
        \nbest_{t}, \pURand_{t}, \nURand_{t}) \\
\label{eq:defvocasexn}
		&= \constriction \bigl[ \pvel_{t} +
			\pcoeff\pURand_{t+1}\otimes(\pbest_{t+1} - \ppos_{t+1}) \\
		& \quad \quad \quad \; + \ncoeff\nURand_{t+1}\otimes(\nbest_{t+1} - \ppos_{t+1})
		\bigr]
\end{align}

>>>>

%% PSO Stuff
\DeclareMathOperator*{\argmin}{arg\;min}
\DeclareMathOperator*{\argmax}{arg\;max}
\DeclareMathOperator*{\arginf}{arg\;inf}
\DeclareMathOperator*{\argsup}{arg\;sup}
\providecommand{\pers}{\ensuremath{P}}
\providecommand{\neigh}{\ensuremath{N}}
\providecommand{\leftind}{\ensuremath{L}}
\providecommand{\rightind}{\ensuremath{R}}
\providecommand{\nURand}{\ensuremath{U^\neigh}}
\providecommand{\pURand}{\ensuremath{U^\pers}}
\providecommand{\ppos}{\ensuremath{\Vec{x}}}
\providecommand{\fppos}{\ensuremath{\Vec{fx}}}
\providecommand{\pvel}{\ensuremath{\Vec{v}}}
\providecommand{\nbest}{\ensuremath{\Vec{x}^\neigh}}
\providecommand{\fnbest}{\ensuremath{\Vec{fx}^\neigh}}
\providecommand{\pbest}{\ensuremath{\Vec{x}^\pers}}
\providecommand{\fpbest}{\ensuremath{\Vec{fx}^\pers}}
\providecommand{\constriction}{\ensuremath{\chi}}
\providecommand{\ncoeff}{\ensuremath{\phi^\neigh}}
\providecommand{\pcoeff}{\ensuremath{\phi^\pers}}
\providecommand{\obs}{\ensuremath{\Vec{\xi}}}
\providecommand{\ofunc}{\ensuremath{f}}
\providecommand{\swarm}{\ensuremath{swarm}}

%SpecExPSO Stuff
\providecommand{\indic}{\ensuremath{I}}
\providecommand{\specvel}{\ensuremath{\vec{V}}}
\providecommand{\specpos}{\ensuremath{\vec{X}}}
\providecommand{\leftn}{\ensuremath{\Vec{x}^\leftind}}
\providecommand{\rightn}{\ensuremath{\Vec{x}^\rightind}}
\providecommand{\caseset}{\ensuremath{\mathcal{C}}}
\providecommand{\casegen}{\ensuremath{c}}
\providecommand{\casedef}{\ensuremath{(\pbest,\nbest)}}
\providecommand{\casexn}{\ensuremath{(S,-)}}
\providecommand{\casexx}{\ensuremath{(S,S)}}
\providecommand{\casexl}{\ensuremath{(S,\leftind)}}
\providecommand{\casexr}{\ensuremath{(S,\rightind)}}
\providecommand{\casepn}{\ensuremath{(-,-)}}
\providecommand{\casepl}{\ensuremath{(-,\leftind)}}
\providecommand{\casepr}{\ensuremath{(-,\rightind)}}
\providecommand{\noeval}[1]{\ensuremath{#1^{-e}}}
\providecommand{\nonbest}[1]{\ensuremath{#1^{-n}}}
\providecommand{\p}{\ensuremath{p}}
\providecommand{\pset}{\ensuremath{\mathbf{p}}}
\providecommand{\s}{\ensuremath{s}}
\providecommand{\sset}{\ensuremath{\mathbf{s}}}
\providecommand{\nsset}{\ensuremath{\mathbf{ns}}}
\providecommand{\n}{\ensuremath{n}}
\providecommand{\nset}{\ensuremath{\mathbf{n}}}
\providecommand{\nnset}{\ensuremath{\mathbf{nn}}}

@@

Particle swarm optimization (PSO) has previously been parallelized primarily by
distributing the computation corresponding to particles across multiple
processors.  In these approaches, additional processors can only add more
particles to the swarm.  However, in many cases this is not efficient when
scaled to very large swarm sizes (on very large clusters).  In this paper we
present a different way of using additional processors in the computation of
PSO motivated by speculative execution in modern computer hardware.  We call
our method Speculative Evaluation in PSO (SEPSO).

In SEPSO future positions of the particles are speculated and evaluated in
parallel with the evaluation of current positions.  We show that a set of
speculative evaluation threads can be constructed such that the set will always
include the position and corresponding evaluation that would normally have
occurred at the next iteration of standard PSO.  Thus SEPSO performs two
iterations of PSO at once, allowing it to achieve better fitness faster.  Not
only can SEPSO can achieve better fitness in the same number of iterations, it
also reaches better fitness for the same number of total function evaluations
in large scale parallel environments.

Speculative Evaluation in PSO can be enhanced by diverging from the behavior of
the standard PSO to make better use of the information obtained by the
speculative evaluations.  Using these algorithmic modifications, we see
dramatic improvements in the effectiveness of the algorithm.

\end{abstract}

@@

Particle swarm optimization (PSO) has been found to be a highly robust and
effective algorithm for solving many types of optimization problems.  For much
of the algorithm's history, PSO was run serially on a single machine.  However,
the world's computing power is increasingly coming from large clusters of
processors.  In order to efficiently utilize these resources for
computationally intensive problems, PSO needs to run in parallel.

@@

These previous parallel techniques distribute the computation needed by the
particles in the swarm over the available processors.  If more processors are
available, these techniques increase the number of particles in the swarm,
either by adding individual particles or by adding entire new sub-swarms, but
in all cases the number of processors never exceeds the number of particles.
The number of iterations of PSO that the algorithms can perform is thus
inherently limited by the time it takes to evaluate the objective
function---additional processors add more particles, but do not make the
iterations go any faster.

For many functions there comes a point of diminishing returns with respect to
adding particles.  In \fig{evals-sphere} we show the value obtained after
20,000 function evaluations (not iterations) as a function of swarm size for
the well-know benchmark function Sphere (20 dimensions, reporting the average
of thirty runs).  Increasing the swarm size from 10 to 50 has a significant
effect on the value obtained.  However, increasing the swarm size from 100 to
200 makes the algorithm less efficient, that is, it reduces the progress the
algorithm makes per evaluation.  For this reason, previous work has recommended
the use of a swarm size of 50 for this function~\cite{bratton-sis07}.  Thus, in
at least some cases, adding particles indefinitely will not yield an efficient
implementation.

@@

In this paper we consider PSO parallelization strategies for clusters of
thousands of processors and functions for which a single evaluation will take
at least tens of seconds, but probably minutes or perhaps hours.  Our purpose
is to explore the question of what to do with a thousand processors when 50 or
100 particles is the recommended swarm size, and simply increasing the swarm
size encounters diminishing returns.

To solve this problem, we propose a method for performing two iterations of PSO
at the same in parallel that we call speculative evaluation.  The name comes
from an analogy to speculative execution techniques commonly used in
processors.  Modern processors, when faced with a branch on which they must
wait (e.g. a memory cache miss), guess which way the branch will go and start
executing, being careful not to make changes that can not be undone.  If the
processor guesses right, execution is much farther ahead than if it had just
idly waited on the memory reference.  If it guesses wrong, execution restarts
where it would have been anyway.  This is called speculative execution or
branch prediction.

%\begin{quote}
%SKIP SKIP
%
%No really, try reading without this! The idea of speculative execution can be applied to some types of optimization
%algorithms.  When the next sampling position of the algorithm does not depend
%on the value of the objective function at the current position,
%future positions can be speculated and evaluated in parallel with the current
%position.
%This is trivially true for optimization by random sampling, for example.
%The next random sample does not depend on the previous point nor on the value
%of the objective function at that point in any way.
%It depends only on a random number.
%There is no need to wait to see
%what objective function value was obtained at the previous point,
%evaluation of the object function at the next point can begin as soon as the next random number is drawn.
%In other optimization algorithms, the next position depends on the previous position, but not the
%value obtained there, a random walk for example.
%In this case
%the next position to be evaluated
%depends only on the previous position and a random number.
%In this case, as it was in the case of random sampling,
%the object function evaluation of the next position can be done in parallel
%because it 
%does not depend on the \emph{value} of the objective function at the previous position.
%There are an infinite number of possible next positions, but only one given the draw of the random number.
%
%In still other cases, the next position to be evaluated may depend on the current position and some
%discrete valued (binary, for example) test on the value obtained at the current position.
%For example, consider a greedy random walk, in which the state only advances if a better
%position is found.
%That is, if the current value is better than the previous value,
%use the sum of the current position and the random number as the next position.
%Otherwise, computer the next position using the sum of the previous position and the random number.
%In this case the objective function can be evaluated at three positions:
%(1) the value of the objective function at the current position,
%(2) the value at the position if the test turns out to be true and (3) the value at the position
%if the test turns out to be false; all in parallel.
%In this example it is important to note that we do not need the value of the objective function at the current
%point to to reduce the infinite number of possible next positions to just two alternatives.
%Choosing between these two alternatives can be done post-hoc, after the objective function value
%at the current position and at the two alternative are computed.
%For all but the most computationally-trivial objective functions, this approach saves almost half of the
%``wall-clock'' time.
%This approach has been applied to simulated annealing~\cite{witte-tpds91} which like the greedy random walk,
%depends on positions and a binary test.
%
%This approach can not be used in gradient methods where the
%step size for sampling typically depends on the gradient and therefor the position and 
%a set of infinitely many possible \emph{values} outcomes
%obtained at the current position (not binary or even finite).
%\end{quote}

@@

In this paper we show that the results of standard PSO can be reproduced
\emph{exactly}, two iterations at a time, using a speculative approach adapted
from speculative execution. We prove that the standard PSO equation can be
factored such that a set of speculative positions can be found which will
\emph{always} include the position computed in the next iteration.  By
computing the value of the objective function for each of the speculative
positions at the same time the algorithm evaluates the objective function for
the current position, it is possible to know the objective function values for
both the current and the next iteration at the same time.  We demonstrate this
principle by implementation and show that it produces exactly the same results
as standard PSO, but two iterations at a time.  The resulting implementation
runs efficiently on large clusters where the number of processors is much
larger than a typical or reasonable number of particles, producing better
results in less ``wall-clock'' time and, surprisingly, with fewer total
function evaluations.

Furthermore, we show that if we relax the requirements of the algorithm, no
longer demanding that it strictly reproduce the exact behavior of standard PSO,
we can introduce new speculative techniques that often out-perform PSO.  These
variants make better use of the information obtained from the extra exploration
made by the speculative function evaluations.  We also explore the idea that,
like branch prediction in processors, we need not speculatively evaluate
\emph{all} possible future positions, we can accelerate the algorithm even if
we are just \emph{likely} to have guessed right.  By pruning the speculation to
just paths that are statistically likely to reproduce the paths that are
equivalent to PSO we can increase the swarm size without increasing the number
of speculative evaluations.  We also consider several recovery strategies for
cases where the pruned set of speculative evaluations does not contain the
evaluation that standard PSO would have done.  A further improvement we explore
is speculating several iterations ahead instead of just one, which is made
possible by pruning the number of speculative evaluations.

@@

At each iteration of the algorithm, the position $\ppos_t$ and velocity
$\pvel_t$ of each particle are updated as follows:
\begin{align}
\nonumber
	\pvel_{t+1} &=
		\constriction \bigl[ \pvel_t
			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
\label{eq:velupdate}
			& \quad \quad \quad \, + \ncoeff\nURand_{t}\otimes(\nbest_{t} - \ppos_{t})
		\bigr] \\
\label{eq:posupdate}
	\ppos_{t+1} &= \ppos_{t} + \pvel_{t+1}
\end{align}
where \( \nURand_{t} \) and \( \nURand_{t} \) are vectors of independent random
numbers drawn from a standard uniform distribution, the \( \otimes \) operator
is an element-wise vector multiplication, $\pbest$ (called personal best) is
the best position the current particle has seen, and $\nbest$ (called
neighborhood best) is the best position the neighbors of the current particle
have seen~\cite{bratton-sis07}.  \( \ncoeff \), \( \pcoeff \), and \(
\constriction \) are parameters with prescribed values required to ensure
convergence (2.05, 2.05, and .73, respectively)~\cite{clerc-tec02}.

There are many ways of defining the neighbors of a particle.  Changing the
neighborhood topology has a significant effect on the performance of the
algorithm and this effect varies widely by function type.  The two most common
topologies used in the literature are the Ring topology and the Complete
topology~\cite{bratton-sis07}.  In the Ring topology, each particle has one (or
more) neighbors to either side of it; in the Complete topology, every particle
is a neighbor to every other particle.  It is generally assumed that, a
particle is also a neighbor to itself in that its own position and value is
considered when updating the particle's personal best, $\nbest$.  Thus with $p$
particles, using the Ring topology each particle has three neighbors: $i-1$,
$i$ (itself), and $i+1$.  With the Complete topology, each particle has $p$
neighbors.

In this paper we use these topologies as well as a parallel adaptation of the
Complete topology, called Random, that has been shown to simulate the behavior
of Complete with far less communication~\cite{mcnabb-cec09}.  In the Random
topology, each particle randomly picks two other particles to share information
with at each iteration, along with itself.  Thus in both the Ring and the
Random topologies, all particles have 3 neighbors.

@@

The PSO algorithm can be trivially parallelized by distributing the processing
needed for each particle on across processors.  But as we have seen in the
introduction, for some functions, and for large numbers of processors, just
adding particles reaches a point of diminishing returns.  That is, adding
processors does not help us reach any given level of fitness appreciably
faster.

To address this issue we present here a different use of additional processors.
Instead of adding particles we employ a speculative evaluation approach as
described in the introduction.  This allows us to perform iterations two at a
time.

To do so we must first refactor PSO such that the determination of the social
and cognitive components, $\nbest$ and $\pbest$, which depend upon the value of
the objective function, is separate from the rest of the computation.  For
simplicity in this explanation, we restrict our discussion here to the case of
a Ring topology with two neighbors which we will call the ``right neighbor''
and ``left neighbor,'' whose positions are represented as $\rightn$ and
$\leftn$ respectively.  Though we will only describe the case of the Ring
topology, our methods can easily be extended to arbitrary topologies.

Our refactoring hinges on the idea that there is a finite set of possible
updates to $\nbest$ and $\pbest$ at each iteration.  For the Ring topology we
are discussing, there are 7 possible update cases, identified in
Table~\ref{tab:evals}.  We label each case with an identifier referring to the
source of the update; a hyphen represents no update, $L$ represents an update
to $\nbest$ coming from the left neighbor, $R$ represents an update to $\nbest$
coming from the right neighbor, and $S$ represents an update to either $\pbest$
or $\nbest$ coming from the particle itself.  In the equations that follow, we
refer to an unspecified update case as $\casegen$, and to the set of cases
collectively as $\caseset$

\begin{table}
  \caption{All possible updates for a particle with two neighbors}
  \label{tab:evals}
  \centering
  \begin{tabular}{lcc}
	Identifier&Source of $\pbest$ update&Source of $\nbest$ update\\
	\hline
	\hline
	$\casepn$&No update&No update\\
	\hline
	$\casepl$&No update&Left Neighbor\\
	\hline
	$\casepr$&No update&Right Neighbor\\
	\hline
	$\casexn$&Self&No update\\
	\hline
	$\casexl$&Self&Left Neighbor\\
	\hline
	$\casexr$&Self&Right Neighbor\\
	\hline
	$\casexx$&Self&Self\\
	\hline
  \end{tabular}
\end{table}

Next, note that the computation of $\pbest$ and $\nbest$ (and hence which case
in $\caseset$ actually occurs) is not seen in the update
equations~\eqref{eq:velupdate} and \eqref{eq:posupdate}.  This update is
usually implemented as a pair of independent if-statements, but that is a bit
awkward for our purposes.  We will instead make this mathematically explicit by
introducing a set of indicator functions, each representing a case from
Table~\ref{tab:evals}.  We can then sum over all of the cases, and the
indicator function will make all of the terms drop to zero except for the case
that actually occurs.

In what follows we will examine the indicator function and the position and
velocity update equations for a specific case, $\casexn$, as presenting
equations for every case would be tedious.  The extension to cases not
presented is straightforward.  When we are performing function minimization, we
can define the indicator function representing the case $\casexn$ by:

\begin{align}
  \nonumber
	\indic_{t+1}^{\casexn} & (\ofunc ( \ppos_{t} ) ,\ofunc(\leftn_{t}),
	\ofunc(\rightn_{t}) ,\ofunc(\pbest_{t-1}) ,\ofunc(\nbest_{t-1}))= \\
  \label{eq:deficasexn}
	&\begin{cases}
	   1 & \text{if} \ \ofunc(\ppos_{t}) < \ofunc(\pbest_{t-1}) \\
	   &\quad \wedge \quad \ofunc(\nbest_{t-1}) < \ofunc(\ppos_{t}) \\
	   &\quad \wedge \quad \ofunc(\nbest_{t-1}) < \ofunc(\leftn_{t}) \\
	   &\quad \wedge \quad \ofunc(\nbest_{t-1}) < \ofunc(\rightn_{t}) \\
	   0 & \text{otherwise}
	\end{cases}
\end{align}

For compactness in our equations we represent this without the parameters since
they can be inferred from the subscripts on $\indic_{t+1}^{\casexn}$.

We now provide notation for the velocity update equation.  Once again we use
$\casexn$ as an example:

\begin{align}
\nonumber
	\specvel_{t+1}^{\casexn} & (\pvel_t, \ppos_{t}, \leftn_{t}, \rightn_{t},
	\pbest_{t-1}, \nbest_{t-1}, \pURand_{t}, \nURand_{t}) \\
\nonumber
		&= \constriction \bigl[ \pvel_{t} +
			\pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
\nonumber
		& \quad \quad \quad \; + \ncoeff\nURand_{t}\otimes(\nbest_{t} -
		\ppos_{t}) \bigr] \\
\nonumber
		&= \constriction \bigl[ \pvel_{t} +
			\pcoeff\pURand_{t}\otimes(\ppos_{t} - \ppos_{t}) \\
\label{eq:defvcasexn}
			& \quad \quad \quad \; + \ncoeff\nURand_{t}\otimes(\nbest_{t-1} -
			\ppos_{t}) \bigr]
\end{align}

Note that we have substituted the actual values obtained for $\pbest$ and
$\nbest$ into Equation~\eqref{eq:defvcasexn}.  Since we have restricted
ourselves to a specific case, we know the values of $\pbest$ and $\nbest$ and
do not need those terms as parameters to $\specvel_{t+1}^{\casexn}$.  In this
case, $\pbest_{t}=\ppos_{t}$, as \pbest was updated by the particle's current
position, and $\nbest_{t}=\nbest_{t-1}$, as \nbest was not updated at iteration
$t$.  Again we will drop the parameters to $\specvel_{t+1}^{\casexn}$ for
compactness.

In the same way we can create notation for the position update
equation~\eqref{eq:velupdate}:
\begin{align}
\label{eq:defpcasexn}
	\specpos_{t+1}^{\casexn} & (\ppos_{t}, \pvel_{t}, \leftn_{t},
	\rightn_{t} ,\pbest_{t-1} ,\nbest_{t-1}, \pURand_{t}, \nURand_{t}) \\
\nonumber
	& = \ppos_{t} + \specvel_{t+1}^{\casexn}
\end{align}
